Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-10 paper/section-general-framework.tex

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple. %: either an index expression
%is a variable that appears in the context (rule \TirName{IVar}), or it
%is an application of an index operation taken from $\IndexOpSet$ to
%other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use
We say that a type is \emph{quantifier-free} if it does not contain
ether universal or existential quantification.

@@

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$, where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. %Note that we have implicity used the fact that
%well-sortedness of index expressions is preserved by addition of extra
%items to the context.
%
Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
\isDefinedAs \sigma(i)$, and on operation symbols as
$\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  %Similarly, 
Given
%a well-indexed type 
$\Delta' \vdash A \isType$, we 
%can apply $\sigma$ to $A$ to produce a new well-indexed type 
we have $\Delta \vdash \sigma^*A
\isType$. The key clauses defining $\sigma^*A$ are for primitive types
and the universal and existential quantifiers:
\begin{displaymath}
  \begin{array}{c}
    \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
    \\
    \begin{array}{c@{\hspace{2em}}c}
      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
      &
      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
    \end{array}
  \end{array}
\end{displaymath}
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
%     \isType$.
%   \end{enumerate}
% \end{lemma}
The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
%$\Delta$: 
$\id_\Delta = (i_1,...,i_n)$ where $\Delta =
i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of 
%two simultaneous substitutions 
$\Delta \vdash \sigma \Rightarrow \Delta'$
and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
(e'_1,...,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
\isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.
%
Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\pi_{i\mathord:s} : \Delta,i\mathord:s \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. %The subscript on
%$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.

@@

A relational environment for a context $\Delta$ is a %(dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
%Every relation environment must respect index expression equality: for
%any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
%substitutions $\Delta \vdash \sigma \Rightarrow
%\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
%\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
%\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
%\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
Relation environments must respect index expression equality: for
all $\tyPrimNm{X} \in \PrimTypeSet$,
$\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
%
We write $\mathrm{RelEnv}(\Delta)$ for the set of all relation
environments for a context $\Delta$. Given %a relational environment
$\rho \in \mathrm{RelEnv}(\Delta)$ and %a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta')$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.

As we stated above, index contexts $\Delta$ are interpreted as certain
%sets of relation environments, i.e.,~
subsets of $\mathrm{RelEnv}(\Delta)$. Depending on the primitive
operations that we assume for our system, we may have different sets
of relational environments that enforce different invariants. 
% We
% require that the sets of relational environments are \emph{valid} in
% the following sense:

@@

% The second condition may
% seem mysterious at first, but it is essential for proving that the
% relational interpretation of universal quantification types behaves
% correctly with respect to application of simultaneous substitution.
%
Given a relation environment $\rho \in \relEnv{E}(\Delta)$, %we define
the set of extensions %$\extends{\rho}{i\mathord:s}$ 
of $\rho$ by %an
%additional index variable 
$i\mathord:s$ %to be 
is $\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi_{i\mathord:s} =
\rho \}$. The set of extensions of a relational environment will be
used in interpretating the universal and existential type formers.  We
assign a relational interpretation to all well-indexed types $\Delta
\vdash A \isType$ by induction on their derivations, parameterised by
%relational environments 
$\rho \in \relEnv{E}(\Delta)$:
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \} \\
  \rsem{\exists i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcup\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following %three
constructions on binary relations: if $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is %defined as 
$\{(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$, and %. With the same assumptions on $R$ and $S$, the relation 
$R
\relTimes S \in \Rel(X \times Y)$ is %defined as 
$\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$, and %. Finally, the relation 
$R \relSum S \in \Rel(X + Y)$ is %defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

@@

We now present the rules for well-typed programs over the 
%collection of 
types %we defined 
in \autoref{sec:algebraically-indexed-types}. Each
well-type program is assigned an index-erasure semantics, building on
the index-erasure semantics of types %we defined 
in \autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus that instantiates our general framework.

@@

We assign an index-erasure semantics to any well-indexed typing
context $\Delta \vdash \Gamma \isCtxt$ by induction:
$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
function $\tmSem{M} : \ctxtSem{\Gamma} \to \tySem{A}$ that completely
ignores the indexing information. In light of
\lemref{lem:tyeqsubst-erasure}, %we define this function is defined
directly on the syntax of well-typed programs, rather than on typing
derivations. The definition of $\tmSem{M}$ is completely standard,
except for the clauses for universal and existential types:
\begin{displaymath}
  \begin{array}{@{}l}
    \begin{array}{@{}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{\hspace{2em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{}}
      \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta
      &
      \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta \\
      \tmSem{\langle [e], M\rangle}\eta & \isDefinedAs & \tmSem{M}\eta
    \end{array} \\
    \tmSem{\mathrm{let}\langle[i], x\rangle = M\ \mathrm{in}\ N}\eta \isDefinedAs \tmSem{N}(\eta, \tmSem{M}\eta)
  \end{array}
\end{displaymath}

We use our index-erasure semantics to define when a pair of programs
are contextually equivalent with respect to syntactically defined
applicative contexts.

@@

  Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
  and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
  \emph{contextually equivalent} ($\Delta; \Gamma_{\mathit{ops}},
  \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$) if for all
  programs $-;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
  \to A) \to \tyPrimNm{unit} + \tyPrimNm{unit}$, it is the case that
  $\tmSem{T\ (\Lambda \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}}
  = \tmSem{T\ (\Lambda \Delta. \lambda
    \Gamma. M_2)}\eta_{\mathit{ops}}$.
\end{definition}

@@

We now state the abstraction theorem for well-typed programs. To state
and prove this theorem for open programs, we extend the
relational interpretation of types to typing contexts. The relational
interpretation of contexts is defined by: % the following clauses: %, where
%we have used the relational lifting of cartesian product $-\relTimes-$
%defined in \autoref{sec:relational-semantics}.
\begin{displaymath}
  \begin{array}{@{}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l@{\hspace{1.5em}}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l}
    \rsem{\epsilon}{\relEnv{E}}{\rho} & \{(*,*)\} &
    \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
  \end{array}
\end{displaymath}
% The relational interpretation of contexts inherits from the relational
% interpretation of types the property of interpreting the application
% of simultaneous substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
  \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
\end{lemma}

@@

We give two constructions of families of sets of relational
environments that allow us to derive useful consequences of
\thmref{thm:abstraction}. To satisfy the hypothesis of
\thmref{thm:abstraction} we still need to prove, for each
instantiation of our general framework, that the program-level
primitive operations preserve all the relations in our
constructions. This is the content of Lemmas
\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
and \ref{lem:metric-environments}.

\paragraph{Compositional Relational Environments}
Our compositional relational environments are defined in terms of
models. To each sort $s \in \SortSet$ we assign a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
$\IndexAxiomSet$.

>>>>

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}
%Given the above data, 
The rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple. %: either an index expression
%is a variable that appears in the context (rule \TirName{IVar}), or it
%is an application of an index operation taken from $\IndexOpSet$ to
%other index expressions (rule \TirName{IOp}). 
The rules for well-indexed types include the usual ones
%rules 
for
%constructing types of 
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). 
We use $\tyPrimNm{bool}$ as an abbreviation for $\tyPrimNm{unit} + \tyPrimNm{unit}$.
The rule \TirName{TyPrim} %allows us to 
forms, from a primitive type $\tyPrimNm{X}$ and appropriately sorted
index expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} 
%permits the formation of 
forms universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types, %are constructed 
formed using the \TirName{TyEx} rule,
%and 
allow for abstraction by hiding. %FIXME: think about the
%introduction of existentials more, perhaps forward ref to their use
We say that a type is \emph{quantifier-free} if it does not contain
ether universal or existential quantification.

@@

For a simultaneous substitution $\Delta \vdash \sigma \Rightarrow
\Delta'$, where $\Delta' = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
either $\Delta$ or $\Delta'$, we can form the \emph{lifted}
simultaneous substitution $\Delta,i\mathord:s \vdash
\sigma_{i\mathord:s} = (\sigma(i_1), ..., \sigma(i_n), i) \Rightarrow
\Delta',i\mathord:s$. %Note that we have implicity used the fact that
%well-sortedness of index expressions is preserved by addition of extra
%items to the context.
%
Application of a simultaneous substitution $\Delta \vdash \sigma
\Rightarrow \Delta'$ to a well-sorted index expression $\Delta' \vdash
e : s$ yields a well-sorted index expression $\Delta \vdash \sigma^*e
: s$. The expression $\sigma^*e$ is defined on variables as $\sigma^*i
\isDefinedAs \sigma(i)$, and on operation symbols as
$\sigma^*(\indexOp{f}(e_1,...,e_n)) \isDefinedAs
\indexOp{f}(\sigma^*e_1, ..., \sigma^*e_n)$.  %Similarly, 
Given
%a well-indexed type 
$\Delta' \vdash A \isType$, 
%can apply $\sigma$ to $A$ to produce a new well-indexed type 
we have $\Delta \vdash \sigma^*A
\isType$. The key clauses defining $\sigma^*A$ are for primitive types
and the universal and existential quantifiers:
\begin{displaymath}
  \begin{array}{c}
    \sigma^*(\tyPrim{X}{e_1,...,e_n}) \isDefinedAs \tyPrim{X}{\sigma^*e_1,...,\sigma^*e_n}
    \\
    \begin{array}{c@{\hspace{2em}}c}
      \sigma^*(\forall i\mathord:s.A) \isDefinedAs \forall i\mathord:s.\sigma_{i\mathord:s}^*A
      &
      \sigma^*(\exists i\mathord:s.A) \isDefinedAs \exists i\mathord:S. \sigma_{i\mathord:s}^*A
    \end{array}
  \end{array}
\end{displaymath}
% \begin{lemma}
%   Let $\Delta \vdash \sigma \Rightarrow \Delta'$ be a simultaneous
%   substitution.
%   \begin{enumerate}
%   \item If $\Delta' \vdash e : s$, then $\Delta \vdash \sigma^*e : s$; and
%   \item If $\Delta' \vdash A \isType$, then $\Delta \vdash \sigma^*A
%     \isType$.
%   \end{enumerate}
% \end{lemma}
The \emph{identity} simultaneous substitution $\Delta \vdash
\id_\Delta \Rightarrow \Delta$ is %just the sequence of variables in
%$\Delta$: 
$\id_\Delta = (i_1,...,i_n)$ where $\Delta =
i_1\mathord:s_1,...,i_n\mathord:s_n$. The \emph{composition} of 
%two simultaneous substitutions 
$\Delta \vdash \sigma \Rightarrow \Delta'$
and $\Delta' \vdash \sigma' \Rightarrow \Delta''$, where $\sigma' =
(e'_1,...,e'_n)$, is defined as $\Delta \vdash \sigma' \circ \sigma
\isDefinedAs (\sigma^*e'_1, ..., \sigma^*e'_n) \Rightarrow \Delta''$.
%
Given a context $\Delta = i_1\mathord:s_1,...,i_n\mathord:s_n$, and a
variable/sort pair $i\mathord:s$ such that $i$ does not appear in
$\Delta$, we define the \emph{projection} simultaneous substitution
$\Delta, i\mathord:s \vdash \pi_{i\mathord:s} \Rightarrow \Delta$ as
$\pi_{i\mathord:s} = (i_1,...,i_n)$. %The subscript on
%$\pi_{i\mathord:s}$ is the variable/sort pair that is being discarded.

@@

A relational environment for a context $\Delta$ is a %(dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
%Every relation environment must respect index expression equality: for
%any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
%substitutions $\Delta \vdash \sigma \Rightarrow
%\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
%\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
%\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
%\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
Relational environments must respect index expression equality: for
all $\tyPrimNm{X} \in \PrimTypeSet$,
$\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$, and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.
%
We write $\mathrm{RelEnv}(\Delta)$ for the set of all relational
environments for a context $\Delta$. Given %a relational environment
$\rho \in \mathrm{RelEnv}(\Delta)$ and %a simultaneous substitution
$\Delta \vdash \sigma \Rightarrow \Delta'$, we can derive the composed
relational environment (with an abuse of notation) $\rho \circ \sigma
\in \mathrm{RelEnv}(\Delta')$ as $(\rho \circ \sigma)\ \tyPrimNm{X}\
\sigma' = \rho\ \tyPrimNm{X}\ (\sigma' \circ \sigma)$.

We interpret index contexts $\Delta$ as certain
%sets of relation environments, i.e.,~
subsets of $\mathrm{RelEnv}(\Delta)$. Depending on the primitive
operations that we assume for our system, we may have different sets
of relational environments that enforce different invariants. 
% We
% require that the sets of relational environments are \emph{valid} in
% the following sense:

@@

% The second condition may
% seem mysterious at first, but it is essential for proving that the
% relational interpretation of universal quantification types behaves
% correctly with respect to application of simultaneous substitution.
%
Given a relational environment $\rho \in \relEnv{E}(\Delta)$, %we define
the set of extensions %$\extends{\rho}{i\mathord:s}$ 
of $\rho$ by %an
%additional index variable 
$i\mathord:s$ %to be 
is $\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi_{i\mathord:s} =
\rho \}$. The set of extensions of a relational environment will be
used in interpreting the universal and existential type formers.  We
assign a relational interpretation to all well-indexed types $\Delta
\vdash A \isType$ by induction on their derivations, parameterised by
%relational environments 
$\rho \in \relEnv{E}(\Delta)$:
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & \isDefinedAs & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e_1,...,e_n}}{\relEnv{E}}\rho & \isDefinedAs & \rho\ {\tyPrimNm{X}}\ (e_1,...,e_n) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & \isDefinedAs & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcap\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \} \\
  \rsem{\exists i\mathord:s.A}{\relEnv{E}}\rho & \isDefinedAs & \bigcup\{ \rsem{A}{\relEnv{E}}{\rho'} \sepbar \rho' \in \extends{\rho}{i\mathord:s} \}
\end{eqnarray*}
In this definition we have made use of the following %three
constructions on binary relations: if $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is %defined as 
$\{(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$, and %. With the same assumptions on $R$ and $S$, the relation 
$R
\relTimes S \in \Rel(X \times Y)$ is %defined as 
$\{((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$, and %. Finally, the relation 
$R \relSum S \in \Rel(X + Y)$ is %defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

@@

We now present the rules for well-typed programs over the 
%collection of 
types %we defined 
in \autoref{sec:algebraically-indexed-types}. Each
well-type program is assigned an index-erasure semantics, building on
the index-erasure semantics of types %we defined 
in \autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics of
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
instantiation of our general framework.

@@

We assign an index-erasure semantics to any well-indexed typing
context $\Delta \vdash \Gamma \isCtxt$ by induction:
$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
function $\tmSem{M} : \ctxtSem{\Gamma} \to \tySem{A}$ that completely
ignores the indexing information. In light of
\lemref{lem:tyeqsubst-erasure}, we do this %we define this function is defined
directly on the syntax of well-typed programs, rather than on typing
derivations. The definition of $\tmSem{M}$ is completely standard,
except for the clauses for universal and existential types:
\begin{displaymath}
  \begin{array}{@{}l}
    \begin{array}{@{}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{\hspace{2em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l@{}}
      \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta
      &
      \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta \\
      \tmSem{\langle [e], M\rangle}\eta & \isDefinedAs & \tmSem{M}\eta
    \end{array} \\
    \tmSem{\mathrm{let}\langle[i], x\rangle = M\ \mathrm{in}\ N}\eta \isDefinedAs \tmSem{N}(\eta, \tmSem{M}\eta)
  \end{array}
\end{displaymath}

We use our index-erasure semantics to define when a pair of programs
are contextually equivalent with respect to syntactically defined
contexts, following Hofmann \cite{hofmann08correctness}.

@@

  Two programs $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_1 : A$
  and $\Delta; \Gamma_{\mathit{ops}}, \Gamma \vdash M_2 : A$ are
  \emph{contextually equivalent} ($\Delta; \Gamma_{\mathit{ops}},
  \Gamma \vdash M_1 \stackrel{ctx}\approx M_2 : A$) if for all
  programs $-;\Gamma_{\mathit{ops}} \vdash T : (\forall \Delta. \Gamma
  \to A) \to \tyPrimNm{bool}$, it is the case that $\tmSem{T\ (\Lambda
    \Delta. \lambda \Gamma. M_1)}\eta_{\mathit{ops}} = \tmSem{T\
    (\Lambda \Delta. \lambda \Gamma. M_2)}\eta_{\mathit{ops}}$.
\end{definition}

@@

We now state the abstraction theorem for well-typed programs. To state
and prove this theorem for open programs, we extend the
relational interpretation of types to typing contexts. The relational
interpretation of contexts is defined by: % the following clauses: %, where
%we have used the relational lifting of cartesian product $-\relTimes-$
%defined in \autoref{sec:relational-semantics}.
\begin{displaymath}
  \begin{array}{@{}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l@{\hspace{1.5em}}l@{\hspace{0.1em}\isDefinedAs\hspace{0.1em}}l}
    \rsem{\epsilon}{\relEnv{E}}{\rho} & \{(*,*)\} &
    \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
  \end{array}
\end{displaymath}
% The relational interpretation of contexts inherits from the relational
% interpretation of types the property of interpreting the application
% of simultaneous substitutions as composition:
% \begin{lemma}\label{lem:ctxtsubst-rel}
%   If $\Delta' \vdash \Gamma \isCtxt$ and $\Delta \vdash \sigma
%   \Rightarrow \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
%   $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
%   \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma)}$.
% \end{lemma}

@@

We give two constructions of families $\relEnv{E}$ of sets of
relational environments that allow us to derive useful consequences of
\thmref{thm:abstraction}. To satisfy the hypothesis of
\thmref{thm:abstraction} we still need to prove, for each
instantiation of our general framework, that the program-level
primitive operations preserve all the relations in our
constructions. This is the content of Lemmas
\ref{lem:geom-environments-1}, \ref{lem:geom-environments-2},
\ref{lem:monoid-ops-related}, \ref{lem:environments-information-flow}
and \ref{lem:metric-environments}.

\paragraph{Model-based Relational Environments}
Our model-based relational environments are defined in terms of
models. To each sort $s \in \SortSet$ we assign a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ satisfying all the axioms in
$\IndexAxiomSet$.
